---
title: Reference
slug: reference
output:
  blogdown::html_page:
    template: ../pandoc/toc-title_html.template
    toc: true
---

```{r include=FALSE}
library(tidyverse)
library(emo)
library(here)

source(here("lib", "graphics.R"))
```


# Citations and bibliography

[^bibtex] You can [download a BibTeX file](/bib/references.bib) of all the non-web-based readings in the course.

[^bibtex]: {-} You can open the file in [BibDesk on macOS](http://bibdesk.sourceforge.net/), [JabRef on Windows](http://www.jabref.org/), or [Zotero](https://www.zotero.org/) or [Mendeley](https://www.mendeley.com/) online.


# Things you should know for Exam 1

## Capitalism

**You should understand…**

- …the components of the capitalist economic system: private property, markets, and firms
- …what happens when any of these components gets distorted
- …what makes public goods different from regular goods (see public goods game)
- …what institutions are and how they coordinate action
- …what GDP is, what it measures, what it doesn't measure, what problems there are with it, what alternatives there are for it, and why it continues to be popular
- …the difference between real and nominal values (and why we care)
- …what a price index is
- …what purchasing power parity (PPP) is (and also what the Big Mac Index is)
- …the downsides of capitalism (inequality + environmental damage)

**Games and simulations played in class:**

- *Pit market trading with paper clips*: how supply and demand are derived from producer willingness to accept and consumer willingness to pay

**Important fomulas:**

- Adjusting for inflation:

    $$
    \text{Real} = \frac{\text{Nominal}}{\text{Price Index / 100}}
    $$

- Percent change:

    $$
    \frac{\text{Current} - \text{Previous}}{\text{Previous}}
    $$

- Compound annual growth rate (CAGR); periodic method (this assumes interest is compounded once a year; this is the harder method and you don't really need to use it):

    $$
    r = exp(\frac{ln(\frac{\text{Price index}_{\text{new}}}{\text{Price index}_{\text{old}}})}{t}) - 1
    $$
- Compound annual growth rate (CAGR): continuous method (this assumes interest is compounded continuously; this is the easier method and you should generally use this):

    $$
    r = \frac{ln(\frac{\text{Price index}_{\text{new}}}{\text{Price index}_{\text{old}}})}{t}
    $$

**Helpful resources:**

- [<i class="fas fa-external-link-square-alt"></i> CPI calculator](https://www.bls.gov/data/inflation_calculator.htm)

\

## Social interactions, economic outcomes, behavioral economics, and incentives

**You should understand…**

- …that perfectly rational individual behavior can create irrational and inferior social outcomes
- …how to use game theory to analyze social interactions. In particular, you should be able to define the following: game, zero-sum, Pareto efficiency, Nash equilibrium, pure strategy, mixed strategy, dominant strategy
- …what social dilemmas, collective action problems, and tragedies of the commons are
- …the difference between a stag hunt game and a prisoners dilemma game, why that difference is important, and why stag hunts are possibly a better metaphor for social dilemmas
- …what preference falsification and why it matters for public policy and administration
- …what factors prevent individuals from cooperating, such as uneven payoffs, lack of assurance, preference falsification, dishonesty, and selfishness
- …how to fix collective action problems with altruism, repetition and iteration, infinitization, punishment, norms, and institutions
- …how heuristics and mental shortcuts go against the assumptions of models based on *homo economicus*
- …the differences between the major categories of heuristics, including representativeness, availability, adjustment and anchoring, and time-based issues such as hyperbolic discounting and interpersonal strategic conflict
- …what a nudge is and how it relates to choice architecture, libertarian paternalism, and public policy and administration
- …how incentives can get crowded out and distorted when extrinsic rewards or punishments replace intrinsic motivation (i.e. don't marketize important social relationships; pay enough or don't pay at all)

**Games and simulations played in class:**

- *Public goods game with black and red cards*: free riding, non-rivalry, and non-excludability
- [*The Evolution of Trust*](http://ncase.me/trust/): prisoners dilemmas and optimal responses and strategies
- Pop quiz 1: Sarah loves to listen to New Age music and faithfully reads her horoscope each day. In her spare time, she enjoys aromatherapy and participating in a local spirituality community.
    - Sarah is likely a holistic healer
    - Sarah is likely a school teacher (correct answer; insensitivity to prior probability of outcomes + representativeness)
- Pop quiz 2: Provo is served by two hospitals. In the larger hospital, about 45 babies are born each day, and in the smaller hospital about 15 babies are born each day. As you know, about 50% of all babies are boys, but the exact percentage varies from day to day—sometimes it's a bit higher than 50%, and sometimes it's lower. The hospitals make special note of when more than 60% of the babies born that day are boys. Which hospital do you thing recorded more of these boy-heavy days?
    - The smaller hospital (correct answer; insensitivity to sample size)
    - The larger hospital
    - About the same
- Pop quiz 3: 
    - Prompt 1: International NGOs don't have regular access to regional international organizations. Today, NGOs participate in only 5% of African Union meetings.
    - Prompt 2: Regional international organizations have become increasingly important. Today, nearly 100% of African countries are part of the African Union.
    - Question: What percent of North American and European countries are members of NATO (the North American Treaty Organizations)? (example of anchoring)
- True random vs. fake random coin flips (misconception of chance)
- $8 \times 7 \times 6 \times 5 \times 4 \times 3 \times 2 \times 1$ vs. $1 \times 2 \times 3 \times 4 \times 5 \times 6 \times 7 \times 8$ (anchoring + estimation based on incomplete information)

**Helpful resources:**

- [<i class="fas fa-external-link-square-alt"></i> Battle of the Sexes](http://policonomics.com/battle-of-the-sexes/) and [mixed strategies](http://policonomics.com/mixed-strategy/)
- [<i class="fas fa-external-link-square-alt"></i> Prisoner's Dilemma and payoffs](http://policonomics.com/video-d3-prisoners-dilemma/)
- [<i class="fab fa-youtube"></i> Chicken](https://www.youtube.com/watch?v=sww-Zsl0IRY)
- [<i class="fas fa-external-link-square-alt"></i> 2×2 game solver to check answers](http://mindyourdecisions.com/GameSolver.html)

\

## Public policy for fairness and efficiency

**You should understand…** 

- …the difference between Pareto efficiency and fairness
- …why Pareto efficiency is not necessarily the best standard for measuring the success of a policy
- …how we can measure fairness with substantive standards, procedural standards, and Rawlsian standards
- …how cultural perceptions of luck and fairness shape public policy
- …how ideas of efficiency and fairness apply to international trade
- …how public policy can be used to change the payoffs in games (e.g. making it more expensive to use water and deplete public goods)
- …what elasticity measures (i.e. what it means for something to be inelastic vs. elastic)
- …why good public policies should be a Nash equilibrium
- …how philosophy and research deisgn allow us to estimate causal relationships
- …how difference-in-difference estimation works

**Games and simulations played in class**

- *Veil of ignorance simulation*: example of how different policies emerge when decided behind a veil of ignorance

\

## Work, wellbeing, and scarcity

**You should understand…**

- …what opportunity costs are and how they influence decision making
- …how to draw a budget line and what budget lines mean
- …how utility is measured and what indifference curves are
- …the difference between the marginal rate of substitution (slope of the indifference curve) and the marginal rate of transformation (slope of the feasible frontier)
- …what it means when marginal product and marginal utility diminish
- …how to find the utility-maximizing level of consumption given preferences and budget constraints
- …the difference between normal and inferior goods
- …what income effects and substitution effects are and how they're related to government policies

**Games and simulations played in class:**

- *XYZ Airlines*: diminishing marginal product

**Important formulas:**

- All the ways marginal utility (or marginal rate of substitution) can be written: 

    $$
    MRS = \frac{dy}{dx} = \frac{\Delta y}{\Delta x} = \frac{\text{Price}_x}{\text{Price}_y} = \frac{MU_x}{MU_y} = \frac{\partial u / \partial x}{\partial u / \partial y}
    $$

**Helpful resources:**

- [<i class="fab fa-youtube"></i> How to draw income and substitution effects](https://www.youtube.com/watch?v=6ObXpWgYbJ8)
- [<i class="fab fa-youtube"></i> Example Income and Subsitution Effects For Normal and Inferior Goods](https://www.youtube.com/watch?v=pLhh_D5b_Lg)
- [<i class="fab fa-youtube"></i> Income and Substitution Effects](https://www.youtube.com/watch?v=GgzS04HCVbw)

\

## Institutions, power, and inequality

**You should understand…**

- …why institutions matter for public administration, policy, and governance
- …the strengths and weaknesses of and the and general differences between three main theoretical approaches to institutions: institutions as rational behavior, institutions as constraints, and institutions as temporary equilibria
- …the difference between informal and formal institutions
- …why informal institutions exert influence over our actions even if they're not officially codified
- …the difference between self-enforcing, self-reinforcing, and self-undermining institutions
- …the role of path dependency in the emergence of institutions
- …how inequality is measured and why we should care about inequality

**Games and simulations played in class**

- *Pólya's Urn*: example of path dependency
- *International trade game*: example of natural institutional emergence, the importance of initial allocations of resources, and the distribution of bargaining power

**Important fomulas:**

- Gini coefficients:

    ```{r gini-example, echo=TRUE}
    # R code
    library(ineq)
    
    # List of incomes
    incomes <- c(10000, 20000, 50000, 100000, 200000)
    
    # Calculate Gini coefficient
    ineq(incomes, type = "Gini")
    ```
    
    ```{r plot-gini, echo=FALSE, fig.width=4, fig.height=3.5}
    incomes_to_plot <- tibble(incomes = c(0, incomes),
                              prop_population = seq(0, 1, 0.2)) %>%
      mutate(prop_income = cumsum(incomes) / sum(incomes))
    
    ggplot(data = incomes_to_plot) +
      geom_ribbon(aes(x = prop_population,
                      ymin = prop_income,
                      ymax = prop_population),
                  fill = "#BF616A", alpha = 0.3) +
      geom_line(aes(x = prop_population, y = prop_income),
                size = 2, color = "#BF616A") +
      geom_abline(slope = 1, intercept = 0, size = 0.5, color = "grey50") +
      labs(x = "Cumulative % of population", y = "Cumulative % of income") +
      scale_x_continuous(expand = c(0, 0), breaks = seq(0, 1, 0.2),
                         labels = scales::percent_format(accuracy = 1)) +
      scale_y_continuous(expand = c(0, 0), breaks = seq(0, 1, 0.2),
                         labels = scales::percent_format(accuracy = 1)) +
      theme_econ(13) +
      theme(panel.grid = element_blank(),
            axis.line = element_line(color = "black", size = 0.25))
    ```

    Alternatively, you can use this formula where $x$ = income, $y$ = cumulative proportion of the population, and $\mu_x$ = mean of income:
    
    $$
    \frac{2}{\mu_x} \text{Cov}(x, y)
    $$

\

# Things you should know for Exam 2

## The firm

**You should understand…**

- …how the decision-making structures of firms and markets are different
- …that perfectly complete contracts are difficult (if not impossible) to create
- …what happens when there are incomplete contracts
- …what a principal-agent problem is
- …adverse selection
- …moral hazard
- …how firms can use the labor discipline model to induce higher worker effort
- …why involuntary unemployment is necessary
- …the conditions under which the outsourcing and privatization of government services can be effective and ethical (or ineffective)

**Games and simulations played in class**

- *Market for lemons*: how adverse selection distorts markets and how the presence of information restores competition to markets

\

## Firms and markets

**You should understand…**

- …how demand curves are derived from consumer willingness to pay
- …the difference between fixed costs and variable costs
- …how to calculate total cost, total revenue, average fixed costs, average variable costs, marginal cost, marginal revenue, and maximum profit
- …that maximum profit occurs where marginal revenue is equal to marginal cost ($MR = MC$)
- …that socially optimal quantity occurs when the demand is equal to the marginal cost ($\text{demand} = MC$)
- …how to calculate elasticity of demand ($-\frac{\Delta Q}{\Delta P} \times \frac{P}{Q}$)
- …what elasticity measures and why it is important in public policy and administration
- …how a single demand curve can have an overall elasticity *and* different elasticities at each point
- …economies of scale, diseconomies of scale, economies of agglomeration, network effects, and the difference between short-run and long-run costs
- …that market equilibria (i.e. optimal price and quantity) occur at the intersection of supply and demand curves
- …how government-imposed price floors and price ceilings distort market-clearing equilibria 
- …and be able to identify the differences between changes in supply/demand and changes in quantity supplied/demanded
- …what consumer and producer surplus represent
- …the relationship between elasticity of supply and/or demand and the size of consumer and producer surplus
- …how taxes impose deadweight loss on society
- …how the burden of taxes depends on the elasticity of supply and/or demand
- …why governments tax and the philosophical and ethical principles behind who should bear the burden of taxes
- …the difference between price-taking and price-making
- …how firms try to gain market power, including monopolies, branding, cost controls, regulation, and switching costs
- …why firms try to gain market power
- …why firms want to price discriminate
- …the consequences of monopolistic production (lower Q and higher P than what would happen under perfect competition; deadweight loss)
- …how governments can regulate monopolies
- …why natural monopolies exist and how governments can induce them to produce at socially optimal levels
- …how firms need to be somewhat anti-competitive and anti-capitalist in order to maximize profits, innovate, and (essentially) be more competitive and capitalist

**Games and simulations played in class:**

- *XYZ Frames*: fixed costs, variable costs, average costs
- *BYU Creamery chocolate milk*: Q, TVC, TFC, TC, MC, AVC, AFC, ATC, and π

**Important fomulas:**

- Demand:

    $$
    P = aQ + b
    $$

- Total cost:

    $$
    TC = TFC + TVC \\
    \text{or} \\
    \text{A formula using } Q \text{, like} \\
    TC = aQ^2 + b
    $$

- Average cost:

    $$
    AC = \frac{TC}{Q}
    $$

- Marginal cost:

    $$
    \begin{aligned}
    MC &= \frac{\Delta TC}{\Delta Q} \\
    &\text{or} \\
    MC &= \text{First derivative of TC} \\
    &= 2aQ \text{ (if } TC = aQ^2 + b)
    \end{aligned}
    $$

- Total revenue:

    $$
    \begin{aligned}
    TR &= PQ \\
    &\text{or} \\
    TR &= (aQ + b)Q \\
    &= aQ^2 + bQ
    \end{aligned}
    $$

- Average revenue:

    $$
    AR = \frac{TR}{Q}
    $$

- Marginal revenue:

    $$
    \begin{aligned}
    MR &= \frac{\Delta TR}{\Delta Q} \\
    &\text{or} \\
    MR &= \text{First derivative of TR} \\
    &= 2aQ + b \text{ (if } TR = aQ^2 + bQ)
    \end{aligned}
    $$

- Maximum profit:

    $$
    max(\pi): MC = MR
    $$

- Price elasticity of demand ([see class 15 for an explanation](/class/15-class/#elasticity-of-demand) of how to get to $- \frac{\Delta Q}{\Delta P} \times \frac{P}{Q}$):

    $$
    \varepsilon = -\frac{\% \text{ change in quantity demand}}{\% \text{ change in price}} = - \frac{\Delta Q}{\Delta P} \times \frac{P}{Q}
    $$

**Important graphs:**

- Consumer surplus, producer surplus, tax revenues, tax burdens, and deadweight loss (use algebra and geometry to figure out the areas of the triangles ($\frac{1}{2} \times b \times h$) and rectangles ($l \times w$)):

    ```{r supply-demand-surplus, echo=FALSE, warning=FALSE, message=FALSE, fig.width=5, fig.height=5}
    source(here("lib", "supply-demand-surplus.R"))
    
    tax_graph(demand, supply, supply_tax, NULL, TRUE)
    ```

**Helpful resources:**

- [<i class="fab fa-youtube"></i> Derivatives of Exponential Functions](https://www.youtube.com/watch?v=xgarjF9DP78) (how to calculate derivatives quickly)
- [<i class="fas fa-file-excel"></i> BYU chocolate milk (empty file)](/files/byu-chocolate-milk.xlsx)
- [<i class="fas fa-file-excel"></i> BYU chocolate milk (completed file)](/files/byu-chocolate-milk-filled.xlsx)
- [<i class="fab fa-youtube"></i> Jason Welker, "Natural Monopoly and the Need for Government Regulation"](https://www.youtube.com/watch?v=LhiyJlI5xa0)

\

## Macroeconomics

**You should understand…**

- …the basic mechanics of fractional reserve banking
- …the role of central banks in overseeing the money supply
- …the Fed's dual mandate and the Fed's three tools for manipulating money supply
- …the difference between monetary policy fiscal policy. Be prepared to identify examples of each type.
- …how federal government budgets differ from state and local government budgets (and household budgets)

**Games and simulations played in class:**

- *Banking in Smallville, USA*: how fractional reserve banking creates money

**Helpful resources:**

- <i class="fab fa-youtube"></i> Super clear explanation of all the Fed's tools for dealing with monetary policy (watch this one!): [What's all the Yellen About? Monetary Policy and the Federal Reserve](https://www.youtube.com/watch?v=1dq7mMort9o)
- <i class="fab fa-youtube"></i> The Atlanta Fed's explanation of monetary policy: [The Fed Explains Monetary Policy](https://www.youtube.com/watch?v=wOfQPn9Jwpo)

\

# Things you should know for Exam 3

## Institutions and public goods problems

**You should understand…**

- …the difference between private goods, club goods, common pool resources, and public goods (and how they can be classified as (non)rivalrous and (non)excludable)
- …what a market failure is
- …what a government failure is
- …what public goods are and how governments, the private sector, and the third sector can address them
- …what common pool resources (CPRs) are and how governments, the private sector, and the third sector can address them
- …the difference between social and private marginal cost (supply) and marginal benefit (demand)
- …what externalities are and how governments, the private sector, and the third sector can address them
- …what Coasian bargaining is, when it is advantageous, and why it sometimes fails
- …how cap and trade systems can fix externalities (and when they can't)
- …how Pigouvian taxes can fix externalities (and when they can't)
- …how regulation can fix externalities (and when and why they can't)^[Remember the peanut butter regulation story—regulation is never strictly economic or mathematical]
- …why not everything should be a market
- …the difference between income and assets
- …why shared national identity and strong horizontal networks of institutions are important for a country's social and economic wellbeing
- …how inequitable public policies lead to decreased public goods provision, unequal institutional access, and increase ethnofractionalization
- …why slavery has had lasting institutional impacts on the economic system of today and how government policies have contributed to these consequences (specifically in housing and education)

**Games and simulations played in class**

- *The paper river*: example of how Coasian bargaining can address externalities without government involvement and why property rights are essential for successful bargaining
