---
title: "Problem set 8"
date: "2019-03-29"
due_date: "2019-03-29"
due_time: "11:59 PM"
editor_options: 
  chunk_output_type: console
output:
  blogdown::html_page:
    css: /css/gt-overrides.css
citeproc: false
bibliography: ../../static/bib/references.bib
csl: ../../static/bib/chicago-fullnote-bibliography-no-bib.csl
---


**Submit this as a PDF on Learning Suite.** You can use whatever you want to make your drawings, including [Gravit Designer](https://gravit.io/), [Desmos](https://www.desmos.com/), Adobe Illustrator, Excel, PowerPoint, Microsoft Paint, or photographed/scanned pen and paper.

**Cite your sources and show your work.**

```{r load-packages, include=FALSE}
library(tidyverse)
library(pander)
library(scales)
library(here)

source(here("lib", "graphics.R"))
```

# 1

The nation of Oskago is a rocky island whose inhabitants are particularly hardworking and adept fishermen. They are able to catch eight tons of fish per year, yet they can gather only four tons of coconuts per year. The going price for one coconut on the island is two fish. Nearby is the island paradise of Silveto whose lazy inhabitants catch only four tons of fish per year but gather eight tons of coconuts every year from the lush forests. On Silveto, the price of one fish is two coconuts. With the advent of lightweight boats, the possibility of virtually cost-free trade now exists. 

Should the two island nations trade products? **Why or why not? Graph your results.**


# 2

In "The Problem of Social Cost" Coase claims that clear assignment of property rights can solve externality problems.   

1. Why does Coase argue that from a social perspective it should not matter whether the polluter pays for pollution or whether those affected by pollution pay the polluter not to pollute? **Answer in ≈30 words.**
2. Under what conditions does the Coase Theorem hold? When might we expect it not provide useful guidance for policy design? Why? **Answer in ≈30 words.**


# 3

Thailand and Cambodia produce rice and trucks at constant rates of product transformation (their possibility frontiers are straight lines—they don't make more or less of the product depending on how much they're currently making). In one year, Thailand could produce 1,500 tons of rice and no trucks, 500 boatloads of trucks and no rice, or any combination in between. Cambodia can produce 1,000 tons of rice or 200 boatloads of trucks, or somewhere in between. **Answer the following**:

1. Draw the production possibility frontier for each country, labeling each curve and known points.
2. Thailand can produce more of both projects than Cambodia. Can Thailand gain anything by trading with Cambodia? **Explain in ≈30 words.**
3. If the answer to #2 is yes, which ratios of trucks to rice create advantages from trade? Which ratios of trucks to rice create disadvantages from trade?


# 4

Construct a table like the one in [Figure 11.14 in ESPP](https://www.core-econ.org/espp/book/text/11.html#figure-11-14) to analyze the possible market failures associated with the decisions below. In each case, identify which markets or contracts are missing or incomplete.

1. You inoculate your child with a costly vaccination against an infectious disease.
2. You use money that you borrow from the bank to invest in a highly risky project.
3. A fishing fleet moves from the overfished coastal waters of its own country to international waters.
4. A city airport increases its number of passenger flights by allowing nighttime departures.
5. You contribute to a Wikipedia page.
6. A government invests in research in nuclear fusion.

```{r market-failure-table, echo=FALSE, results="asis"}
tibble(thing = c("Decision", "How it affects others", "Cost or benefit",
                 "Market failure (misallocation of resources)",
                 "Possible remedies",
                 "Terms applied to this type of market failure"),
       value = c("Vaccination", rep("—", 5))) %>%
  mutate(thing = fct_inorder(thing)) %>% 
  spread(thing, value) %>% 
  pandoc.table(split.table = Inf, justify = "llllll")
```


# 5

What do you think about the following (hypothetical) market exchanges? You may assume in each case that the people involved are sane, rational adults who have thought about the alternatives and consequences of what they are doing. 

In each case, decide (1) whether you approve, (2) whether you think the transaction should be prohibited, and (3) what institutional or normative reasons support your decision. Discuss each in **≈30 words**.

1. Space Marketing Inc. announces plans to launch giant billboards made from Mylar sheets into low orbit. Companies would pay more than \$1 million dollars to display advertisements. Logos, about the size of the moon, will be visible to millions of people on Earth.
2. You are waiting in line to buy tickets for a movie that is almost sold out. Someone from the back of the line approaches the person in front of you and offers her $25 to let him in front of her.
3. A politically apathetic person, who never votes, agrees to vote in an election for the candidate who pays him the highest amount.
4. A well-informed and sane adult, with an adequate income, decides that he would like to sell himself to become the slave of another person. He finds a buyer willing to pay his asking price. The aspiring slave will give the price paid by the buyer to his children to further their education. 


# 6

In two-car automobile accidents, [passengers in the larger vehicle are significantly more likely to survive than passengers in the smaller vehicle](http://www.buffalo.edu/news/releases/2013/05/026.html). Some politicians and lobbyists have argued that this provides a rationale for encouraging the sale of larger vehicles and discouraging legislation that would induce automobile manufacturers to make smaller cars. Evaluate this argument using the concept of externalities. Are these politicians and lobbyists right? **Answer in ≈40 words.**


# 7

Using a nationwide survey, economists have found that the demand for refrigerators is

$$
P = 1500 - 0.75Q
$$

Based on the internal costs of appliance manufacturers, economists have found that the marginal private cost of producing refrigerators is

$$
P = 550 + 0.5Q
$$

A key component of the refrigerator production process is [chlorofluorocarbon (CFC)](https://en.wikipedia.org/wiki/Chlorofluorocarbon), which helps with refrigeration. CFCs cause environmental damage, though, and economists have estimated that each refrigerator prodcued causes social environmental damage of $0.4Q$.^[Super big hint: this will need to be added to the marginal private cost, so you should end up with $P = 550 + 0.9Q$]

1. What is the unregulated market output and market price given the supply and demand?
2. How many refrigerators should be produced to achieve the socially optimal level? What should the market price be for this level of output?
3. How much total damage to society does this overproduction of refrigerators cause?^[Hint: think about deadweight loss.]
4. One way to achieve the socially optimal level is for the government to impose a Pigouvian tax on the producers. How big of a tax should the government impose?
